Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2021-01-06 Nicolas.tex
YC comments

\textbf{We will thus consider a set of cuts that will define a first fiducial
sample, taking the farthest limits possible. But as this solution might be
over simplistic to create a complete sample, e.g., for it ignores follow-up
efficiency, we also consider another set of cuts to define a conservative
sample. This latter is smaller and therefore less statistically
constraining, but also inherently less prone to selection effects. If the
redshift drift is still significant in the conservative sample, it would be
even more significant in a carefully-tailored selection-free sample. The
goal of this study is to build the first implementation of a SN Ia
properties drift modeling before refining it on a more complex, yet more
accurate, follow-up analysis accounting at the same time for precise selection
modeling using, e.g., the SNANA package \citep{SNANA}.}

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2020} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). \textbf{For this dataset, spectroscopic screening was
    done for candidates with $r \lesssim 19.5$. Then, redshift cuts were applied
    when selecting which SN Ia to follow, giving a redshift range of 0.02 < z <
    0.09 ; further reduced to $<0.08$ in \cite{rigault2020} for extracting local
host properties.} These 114 SNfactory SNe~Ia are thus in the volume-limited part
of the survey (Aldering et al., in prep.), and are therefore assumed to be a
random sampling of the underlying SN population. The SNfactory sample is
particularly useful for studying SN property drift, as it enables us to have a
large complete SN~Ia sample at $z<0.1$.

Finally, we include the HST sample from Pantheon \citep{strolger04}.
\textbf{These high-redshift SNe are of great interest as they provide the
    greatest leverage for testing evolution. While at these redshift the
    supernovae typing is challenging, the target classification was robust
enough to include them within the cosmological analysis \citep{scolnic2018a} and
we do not add further quality cuts. Section~\ref{sec:results} highlights that
our results are not dependent on the inclusion of this dataset.}

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{\lim}$ in Fig.~\ref{fig:sample}. \textbf{We observe
that the apparent fraction of low-stretch SNe (typically $x_1 < -1$) seems to
decrease as a function of redshift.} The evolution of the mean stretch is also
shown in Fig.~\ref{fig:modelall}, where the data are split in redshift bins of
regular sample size. We see that SNe~Ia at higher redshift have on average
larger stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than those at lower redshift
($-0.17\pm 0.10$ at $z\sim0.05$), suggesting that the underlying stretch
distribution is \textbf{evolving with redshift}.

@@

\textbf{In section~\ref{ssec:cuts}, we have built a volume-limited sample from a
set of magnitude-limited ones, using a simple redshift limit. This
simplistic approach is statistically sub-optimal, but is enough to
demonstrate our key finding: a redshift evolution of stretch compatible with
the \cite{rigault2020} model. However, there remains the possibility that
complex selection function related to follow-up efficiencies below our fiducial
(or even conservative cuts) might significantly affect our sample, making it not
fully volume-limited, which could, in turn, bias our conclusion on the
astrophysical drift of the SNe~Ia population.}

\textbf{To test for the existence of potentially significant leftover selection
biases in our sample, we compare the stretch and color distributions of the
SNe~Ia in our sample between datasets of overlapping redshifts. If indeed we
have built a volume-limited sample, the distributions of any SNe~Ia
parameter at a given redshift should not depend on their survey of origin. When
considering a redshift range, this has to be narrow enough such that any drift
would be negligible.}

\textbf{The two samples that overlap the most in redshift are PS1 and SDSS in
the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}). This
corresponds to the 146 SNe~Ia from SDSS at the highest redshift end and thus
the most likely to be affected by selection effects not eliminated by our
redshift cut (see the corresponding discussion in section~\ref{ssec:cuts}).
In that same redshift range, PS1 has 52 SNe~Ia, which are in the lowest
redshift bins and thus unlikely to have any selection issue. To identify
potential inconsistency between PS1 and SDSS, Fig.~\ref{fig:distrib}
presents the stretch and color distribution of both these surveys, together
with $p$-values of Kolmogorov-Smirnov (KS) tests. These $p$-values are too
high ($>10\%$) and do not indicate that these subsamples are inconsistent,
in agreement with the visual impression from Fig.~\ref{fig:distrib}.}

\textbf{We perform a similar analysis for PS1 and SNLS in the redshift range
$0.20 < z < 0.31$ where the same conclusion can be drawn, i.e., there is no
sign of discrepancy in the stretch and color distributions between the low-
and high-end of the SNLS and PS1 datasets, respectively. Nonetheless, the
small size of the SNLS dataset at $z < 0.31$ (12 SNe~Ia vs. 90 for PS1)
limits the sensitivity of the test, as only strong deviation would be
noticeable. Extending the redshift range to $0.20 < z < 0.40$ (though we
have no PS1 data above 0.3) enables to get 31 SNe~Ia from SNLS in total, yet
the stretch p-values remains high at 34\% showing no sign of inconsistency}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SDSS_SNLS_PS1.pdf}
    \caption{\textbf{$x_1$ (left) and $c$ (right) distribution histograms of
            different surveys overlapping in redshift. \textit{Facing up}: SDSS
            and PS1 within the redshift range $0.10 < z < 0.20$; \textit{facing
            down}: PS1 and SNLS within the redshift range $0.20 < z < 0.31$.
            Error bars show the Poisson noise. Our stretch ``Base'' model is
            illustrated in orange at the mean redshift of the redshift ranges,
            $0.15$ and $0.25$, respectively. Kolmogorov-Smirnov test $p$-values
            are indicated on the top (resp. bottom) of each panel showing no
            sign that the SDSS and PS1 (resp. PS1 and SNLS) $x_1$ and $c$
            distributions are not drawn from the same underlying
    distributions.}}
    \label{fig:distrib}
\end{figure}

\textbf{We highlight that the SNe~Ia color is more prone to selection effects
than stretch as illustrated in Fig.~\ref{fig:maglim}; see also e.g., Fig.~3
of \cite{kessler2017}. Hence, since the color distribution shows no
significant hint of leftover selection effect, it further supports our claim
that our simple redshift-based selection criteria is sufficient to build the
SNe~Ia sample required to test the redshift evolution of the stretch
distribution.}

@@

\subsection{Base model applied to data}\label{sec:basemodelapplied}

@@

\textbf{In order to ensure that our results are not driven by the poorly-modeled
    HST sub-sample, we recompute $\Delta$AIC for each model excluding this
    dataset; we find (as listed in Table~\ref{tab:comp}) -3.4, -24.5, -27.2,
    -29.2, -26.9, -60.0 and -5.4, -17.4, -20.4, -22.4, -19.7, -57.9 for the
    fiducial and conservative samples, respectively. The consistency of these
    values with those in Table~\ref{tab:comp} show that the HST sub-sample does
not drive our conclusions.}

>>>>

\textbf{We will thus consider a set of cuts that will define a first fiducial
sample, taking the farthest limits possible. \yc{YC: This sounds too strong and ill-defined. I suggest ``taking limits as initially suggested by completeness analysis''.} \yc{However,} as this solution might be
over simplistic to create a complete sample, e.g., for it ignores follow-up
efficiency, we also consider another set of cuts to define a \yc{so-called} conservative
sample. This latter is smaller and therefore less statistically
constraining, but also inherently less prone to selection effects. If the
redshift drift is still significant in the conservative sample, it would be
even more significant \yc{meaningful? (to avoid repetition)} in a carefully-tailored selection-free sample. The
goal of this study is to build \yc{test?} the first implementation of a SN Ia
properties drift modeling \yc{YC: modeling or model?} before refining it on a more complex, yet more
accurate, follow-up analysis accounting at the same time \yc{YC: not clear, "same time" as what?} for precise selection
modeling using, e.g., the SNANA package \citep{SNANA}.}

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2020} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). \textbf{For this dataset, spectroscopic screening was
    done for candidates with $r \lesssim 19.5$; redshift cuts were \yc{then} applied
    when selecting which SN Ia to follow, \yc{resulting in} a redshift range of $0.02 < z <
    0.09$, further reduced to $<0.08$ in \cite{rigault2020} for extracting local
host properties.} These 114 SNfactory SNe~Ia are thus in the volume-limited part
of the survey (Aldering et al., in prep.), and are therefore assumed to be a
random sampling of the underlying SN population. The SNfactory sample is
particularly useful for studying SN property drift, as it enables us to have a
large complete SN~Ia sample at $z<0.1$.

Finally, we include the HST sample from Pantheon \citep{strolger04}.
\textbf{These high-redshift SNe are of great interest as they provide the
    greatest leverage for testing evolution. While at these redshifts the
    supernovae typing is challenging, the target classification was robust
enough to include them within the cosmological analysis \citep{scolnic2018a} and
we do not add further quality cuts. Section~\ref{sec:results} highlights that, \yc{while compatible with it,}
our results are not dependent on the inclusion of this dataset.}

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{\lim}$ in Fig.~\ref{fig:sample}. \textbf{We observe
that the apparent \yc{YC: why "apparent" here? It's supposed to be intrinsic now that we're working on a complete sample. Too much "observe/apparent/seem".} fraction of low-stretch SNe (typically $x_1 < -1$) seems to
decrease as a function of redshift.} The evolution of the mean stretch is also
shown in Fig.~\ref{fig:modelall}, where the data are split in redshift bins of
regular sample size. We see that SNe~Ia at higher redshift have on average
larger stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than those at lower redshift
($-0.17\pm 0.10$ at $z\sim0.05$), suggesting that the underlying stretch
distribution is \textbf{evolving with redshift}.

@@

\textbf{In section~\ref{ssec:cuts}, we have built volume-limited sample\yc{s} from a
set of magnitude-limited ones, using simple redshift \yc{cuts}. This
simplistic approach is statistically sub-optimal, but \yc{should be} enough to
demonstrate our key finding: a redshift evolution of stretch compatible with
\yc{\sout{the}} \cite{rigault2020} model. However, \yc{the possibility remains} that \yc{a}
complex selection function related to follow-up efficiencies below our fiducial
(or even conservative) cuts \yc{could still}  affect our sample, making it not
fully volume-limited; \yc{this would}, in turn, bias our conclusion on the
astrophysical drift of the SNe~Ia population. \yc{We now look at this possibility.}}

\textbf{To test for the existence of \yc{potential} leftover selection
biases in our sample, we compare the stretch and color distributions of the
SNe~Ia \yc{\sout{in our sample} originating from different} datasets \yc{over} overlapping redshift \yc{ranges}: \yc{these distributions should be similar. Note that the} redshift range has to be narrow enough so that any drift
would be negligible.}

\textbf{The two samples that overlap the most in redshift are PS1 and SDSS in
the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}). This
corresponds to the 146 SNe~Ia from SDSS at the \yc{high} redshift end and thus
the most likely to be affected by \yc{residual} selection effects (see the corresponding discussion in section~\ref{ssec:cuts}).
\yc{Over} that same redshift range, PS1 has 52 SNe~Ia, which are in the lowest
redshift bins and thus unlikely to have any selection issue. To identify
potential inconsistency between PS1 and SDSS \yc{subsamples}, Fig.~\ref{fig:distrib}
\yc{(upper panels) compares} the stretch and color distribution of both these surveys. \yc{The associated Kolmogorov-Smirnov (KS) similarity test $p$-values} are too
high ($>10\%$) \yc{to conclude on any inconsistency},
in agreement with the visual impression from Fig.~\ref{fig:distrib}.}

\textbf{We perform a similar analysis for PS1 and SNLS \yc{over} the redshift range
$0.20 < z < 0.31$ \yc{(Fig.~\ref{fig:distrib}, lower panels)}, where the same conclusion can be drawn: there is no \yc{substantial}
sign of discrepancy in the stretch and color distributions between the low-
and high-end of the SNLS and PS1 datasets, respectively. Nonetheless, the
small size of the SNLS dataset at $z < 0.31$ (12 SNe~Ia vs. 90 for PS1)
limits the sensitivity of the test, \yc{and} only strong deviation would be
noticeable. Extending the redshift range to $0.20 < z < 0.40$ (though we
have no PS1 data above 0.3) \yc{allows to increase the SNLS subsample to 31}, yet
the stretch $p$-values remains high (34\%) showing no sign of inconsistency.}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SDSS_SNLS_PS1.pdf}
    \caption{\textbf{$x_1$ (left) and $c$ (right) distribution histograms of
            different surveys overlapping in redshift. \textit{Facing up}: SDSS
            and PS1 within the redshift range $0.10 < z < 0.20$; \textit{facing
            down}: PS1 and SNLS within the redshift range $0.20 < z < 0.31$.
            Error bars show the Poisson noise. Our stretch ``Base'' model is
            illustrated in orange at the mean redshift of the redshift ranges,
            $0.15$ and $0.25$, respectively. Kolmogorov-Smirnov test $p$-values
            are indicated on the top (resp. bottom) of each panel showing no
            sign that the SDSS and PS1 (resp. PS1 and SNLS) $x_1$ and $c$
            distributions are not drawn from the same underlying
    distributions. \yc{YC: I still urge you to make errorbars and text more visible!}}}
    \label{fig:distrib}
\end{figure}

\textbf{We \yc{finally} highlight that the SNe~Ia color is more prone to selection effects
than stretch as illustrated in Fig.~\ref{fig:maglim}; see also e.g., Fig.~3
of \cite{kessler2017}. Hence, since the \yc{comparison of} color distributions shows no
significant hint of leftover selection effect, it further supports our claim
that our simple redshift-based selection criteria is sufficient to build \yc{a complete}
SNe~Ia sample required to test the redshift evolution of the stretch
distribution.}

@@

\subsection{\yc{Comparison} to data}\label{sec:basemodelapplied}

@@

\textbf{In order to ensure that our results are not driven by the poorly-modeled
    HST sub-sample, we recompute $\Delta$AIC for each model excluding this
    dataset; we find (as listed in Table~\ref{tab:comp}) -3.4, -24.5, -27.2,
    -29.2, -26.9, -60.0 and -5.4, -17.4, -20.4, -22.4, -19.7, -57.9 for the
    fiducial and conservative samples, respectively. \yc{YC: IMHO, it would be better to add these values in 2 new dedicated columns. Or are these precise values really needed? Just say something like "does not change $\Delta$AIC by more than few tenths".} The consistency of these
    values with those in Table~\ref{tab:comp} show that the HST sub-sample does
not drive our conclusions.}
